### 2019-05-06 papers/haskell19/5-typelevel.tex
Pass over Section 6

And simply obtain a concrete generator by refering to |Html_Spec| when using
|genRep|:

@@

In order to represent its data constructors, we can follow the same recipe
presented in Section \ref{sec:representation}, but also parametrizing our
representations over the type variable |a| as well:

@@

\begin{code}
data Some (f :: * -> * -> *) (r :: *) = forall (a :: *) dot SomeC (f a r)
\end{code}
%
The type constructor |Some| is a wrapper for a 2-parametric type that hides the
first type variable using an explicit existential quantifier.
%
Note thus that the type parameter |a| does not appears at the left hand side of
|Some| on its defition.
%
Then, when deriving any |Con|, |Fun| or |Pat| type instance, we will wrap each
construction representation using a |Some|:

>>>>

And simply obtain a concrete generator by referring to |Html_Spec| when using
|genRep|:

@@

In order to represent its data constructors, we can follow the same recipe
presented in Section \ref{sec:representation}, but also parameterizing our
representations over the type variable |a| as well:

@@

\begin{code}
data Some (f :: * -> * -> *) (r :: *) = forall (a :: *) dot SomeC (f a r)
\end{code}
%
The type constructor |Some| is a wrapper for a 2-parametric type that hides the
first type variable using an explicit existential quantifier.
%
Note thus that the type parameter |a| does not appears at the left hand side of
|Some| on its definition.
%
Then, when deriving any |Con|, |Fun| or |Pat| type instance, we will wrap each
construction representation using a |Some|:

### 2019-05-06 papers/haskell19/6-casestudies.tex
Pass over Section 6

The random generation framework presented throughout this work allows us to
write extensible generators in a very concise way.
%
However, this expresiveness comes attached to a perceptible runtime overhead,
primarly inherited from the using the Data Types \`a la Carte---a technique
which is not often scrutinized for performance.
%
In this section, we evaluate the implicit cost of composing generators using
three real-world case studies.

As we have shown in Section \ref{sec:generators}, the random generation
process we propose in this paper can be seen as having two phases.
%
First, generate random values from the representation types used to specify the
shape of our data; and then we use their algebras to translate them to the
corresponding values of our target data types.
%
In particular, the tranformation step is expected to pattern match repeatedly
against the |InL| and |InR| constructors of the |oplus| operators when
traversing each construction injection.
%
Because of this, in general, we expect a performance impact with respect to
manually-written concrete generators.

@@

Then, we analyzed the performance of generating random values using three case
studies:
%
\begin{inparaenum}[(i)]
\item Red-Black Trees (RBT), inspired by Okasaki's formulation
  \cite{okasaki1999red},
\item Lisp S-expressions (SExp), inspired by the package
  \emph{hs-zuramaru}\footnote{\url{http://hackage.haskell.org/package/zuramaru}},
  and
\item HTML expressions (HTML), inspired by the \emph{html}
  package\footnote{\url{http://hackage.haskell.org/package/html}}, which follows
  the same structure as our motivating |Html| example.
\end{inparaenum}
%
The expanse of each case study can be summarized as follows:

@@

These case studies provide a good combination of constructions, interface
functions and patterns, and cover from smaller to larger numbers of
constructions.

We benchmarked the execution time required to generate and fully evaluate 10000
random values corresponding to each case study, comparing both manually-written
concrete generators and those obtained using our modular approach.
%
For this purpose, we used the \emph{Criterion} \cite{criterion} benchmarking
tool for Haskell, and limited the maximum depth of the generated values to five
levels.
%
Additionally, our modular generators were tested using both linear and balanced
generation specifications.
%
Figure \ref{fig:times} illustrates the relative execution time of each case
study, normalized to their corresponding manually-written counterparts---we
encourage the reader to obtain a colored version of this work.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}
      [ height=6cm
      , width=9cm
      , ybar
      , ymajorgrids=true
      , ytick={0,1,...,4}
      , ymax=4
      , ymin=0.5
      , enlargelimits=0.25
      , enlarge x limits=0.25
      , /pgf/bar width=17pt
      , legend style={
        at={(0.5,-0.15)},
        anchor=north,legend columns=3}
      , legend image code/.code={%
        \draw[#1, draw=none] (0cm,-0.1cm) rectangle (0.3cm,0.2cm); }
      , ylabel={Relative Generation Time}
      , symbolic x coords={RBT, SExp, HTML}
      , xtick=data
      % , nodes near coords
      , nodes near coords align={vertical}
      ]
      \addplot[fill, red!30]   coordinates {(SExp,1)    (RBT,1)    (HTML,1)};
      \addplot[fill, green!30] coordinates {(SExp,2.40) (RBT,3.57) (HTML,4.27)};
      \addplot[fill, blue!30]  coordinates {(SExp,1.60) (RBT,2.52) (HTML,1.55)};
      \legend{Manual, Modular (lin), Modular (bal)}
    \end{axis}
  \end{tikzpicture}
  \caption{Generation time comparison between concrete and modular generators.}
  \label{fig:times}
\end{figure}

As it can be observed, our approach suffers from a noticeable runtime overhead
when using linearly defined representations, specially when considering the HTML
case study, which spans a large amount of constructions in the generation
process.
%
However, we found that, by balancing our representation types the generation
performance improves dramatically.
%
At the light of these improvements, our tool includes an additional type-level
computation that automatically balances our representations in order to reduce
the generation overhead.

@@

All in all, we consider that these results are fairly encouraging, given that
the flexibility obtained from using our compositional approach does not produce
severe slowdowns in when generating random values in practice.

>>>>

The random generation framework presented throughout this paper allows us to
write extensible generators in a very concise way.
%
However, this expressiveness comes attached to a perceptible runtime overhead,
primarily inherited from the use of Data Types \`a la Carte---a technique which
is not often scrutinized for performance.
%
In this section, we evaluate the implicit cost of composing generators using
three real-world case studies.

As we have shown in Section \ref{sec:generators}, the random generation
process we propose in this paper can be seen as having two phases.
%
First, we generate random values from the representation types used to specify
the shape of our data; and then we use their algebras to translate them to the
corresponding values of our target data types.
%
In particular, the transformation step is expected to pattern match repeatedly
against the |InL| and |InR| constructors of the |oplus| operators when
traversing each construction injection.
%
Because of this, in general, we expect a performance impact with respect to
manually-written concrete generators.

@@

Later, we analyzed the performance of generating random values using three case
studies:
%
\begin{inparaenum}[(i)]
\item Red-Black Trees (RBT), inspired by Okasaki's formulation
  \cite{okasaki1999red},
\item Lisp S-expressions (SExp), inspired by the package
  \emph{hs-zuramaru}\footnote{\url{http://hackage.haskell.org/package/zuramaru}},
  and
\item HTML expressions (HTML), inspired by the \emph{html}
  package\footnote{\url{http://hackage.haskell.org/package/html}}, which follows
  the same structure as our motivating |Html| example.
\end{inparaenum}
%
The magnitude of each case study can be outlined as follows:

@@

These case studies provide a good combination of data constructors, interface
functions and patterns, and cover from smaller to larger numbers of
constructions.

Then, we benchmarked the execution time required to generate and fully evaluate
10000 random values corresponding to each case study, comparing both
manually-written concrete generators, and those obtained using our modular
approach.
%
For this purpose, we used the \emph{Criterion} \cite{criterion} benchmarking
tool for Haskell, and limited the maximum depth of the generated values to five
levels.
%
Additionally, our modular generators were tested using both linear and balanced
generation specifications.
%
Figure \ref{fig:times} illustrates the relative execution time of each case
study, normalized to their corresponding manually-written counterpart---we
encourage the reader to obtain a colored version of this work.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}
      [ height=6cm
      , width=9cm
      , ybar
      , ymajorgrids=true
      , ytick={0,1,...,4}
      , ymax=4
      , ymin=0.5
      , enlargelimits=0.25
      , enlarge x limits=0.25
      , /pgf/bar width=17pt
      , legend style={
        at={(0.5,-0.15)},
        anchor=north,legend columns=3}
      , legend image code/.code={%
        \draw[#1, draw=none] (0cm,-0.1cm) rectangle (0.3cm,0.2cm); }
      , ylabel={Relative Generation Time}
      , symbolic x coords={RBT, SExp, HTML}
      , xtick=data
      % , nodes near coords
      , nodes near coords align={vertical}
      ]
      \addplot[fill, red!30]   coordinates {(SExp,1)    (RBT,1)    (HTML,1)};
      \addplot[fill, green!30] coordinates {(SExp,2.40) (RBT,3.57) (HTML,4.27)};
      \addplot[fill, blue!30]  coordinates {(SExp,1.60) (RBT,2.52) (HTML,1.55)};
      \legend{Manual, Modular (lin), Modular (bal)}
    \end{axis}
  \end{tikzpicture}
  \caption{Generation time comparison between concrete and modular generators.}
  \label{fig:times}
\end{figure}
%
As it can be observed, our approach suffers from a noticeable runtime overhead
when using linearly defined representations, specially when considering the HTML
case study, involving a large number of constructions in the generation process.
%
However, we found that, by balancing our representation types, the generation
performance improves dramatically.
%
At the light of these improvements, our tool includes an additional type-level
computation that automatically balances our representations in order to reduce
the generation overhead.

@@

All in all, we consider that these results are fairly encouraging, given that
the flexibility obtained from using our compositional approach does not produce
severe slowdowns when generating random values in practice.
